---
title: "Malaria Project: Analyzing Malaria Under 5 Cases Per Thousand Against Lagged Weather Data"
author: "Emma Jones"
date: "October 22, 2018"
output: word_document
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)

mal <- read.csv("./MozSyntheticMalaria.csv")
mal <- mal[,c(1:4, 12:26)]

# creating cases per thousand variable
mal$denom <- mal$Population_UN*mal$u5weight
mal$cpt <- (mal$malaria/mal$denom)*1000

head(mal)

```


```{r lag}
# creating a lagged rain total variable k number of weeks behind the rest of the data
mal$lag2rainT <- lag(mal$rainTot, 2)

mal$lag4rainT <- lag(mal$rainTot, 4)

mal$lag8rainT <- lag(mal$rainTot, 8)

mal$lag16rainT <- lag(mal$rainTot, 16)

# creating a lagged temp avg variable
mal$lag2tavg <- lag(mal$tavg, 2)

mal$lag4tavg <- lag(mal$tavg, 4)

mal$lag8tavg <- lag(mal$tavg, 8)

mal$lag16tavg <- lag(mal$tavg, 16)
```

#Questions of Interest
##1. Which lags are most associated with malaria incidence for temperature and total rain-fall?

```{r rainTcpt}
library(colorspace)
library(ggplot2)

# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

cptRain <- ggplot(data=mal) + geom_smooth(mapping = aes(x = Epiweek, y= cpt), color = "#660066") +
  geom_smooth(mapping = aes(x = Epiweek, y = rainTot), color="#660066") +
  geom_smooth(mapping = aes(x = Epiweek, y=lag2rainT), color="#CC0000") + 
  geom_smooth(mapping = aes(x = Epiweek, y=lag4rainT), color="#FF6600") +
  geom_smooth(mapping = aes(x = Epiweek, y=lag8rainT), color="#FF9933") 
cptRain

# use function ccf in stats package

rain.ccf <- ccf(mal$rainTot, mal$cpt)
rain.ccf

rain.temp.ccf <- ccf(mal$rainTot, mal$tavg)
rain.temp.ccf


```

```{r tavgCpt}

cptTavg <- ggplot(data=mal) + geom_smooth(mapping = aes(x = Epiweek, y= cpt), color = "#660066") +
  geom_smooth(mapping = aes(x = Epiweek, y = tavg), color="#660066") +
  geom_smooth(mapping = aes(x = Epiweek, y=lag2tavg), color="#CC0000") + 
  geom_smooth(mapping = aes(x = Epiweek, y=lag4tavg), color="#FF6600") +
  geom_smooth(mapping = aes(x = Epiweek, y=lag8tavg), color="#FF9933") 
cptTavg

temp.ccf <- ccf(mal$tavg, mal$cpt)
temp.ccf


```
It appears that Total Rain lagged by 4 weeks is most correlated with CPT, while Average Temperature lagged by 16 weeks is most correlated with CPT.

```{r overTime}
library(dplyr)
mal <- mutate(mal, hyp.date = as.Date(paste(mal$Epiyear, mal$Epiweek, "1", sep="-"), format="%Y-%U-%u"))

#malLongi <- ggplot(data = mal, aes(x=hyp.date, y=cpt, color=Region)) +  geom_point(alpha=0.05) + geom_smooth(color="black")
#malLongi + facet_wrap(vars(mal$Region))

#rainTLongi <- ggplot(data = mal, aes(x=hyp.date, y=lag4rainT, color=Region)) + geom_point(alpha=0.05) + 
#  geom_smooth(color="black")
#rainTLongi + facet_wrap(vars(mal$Region))

rainCPT <- ggplot(data = mal, aes(x=hyp.date, y=sqrt(lag4rainT), color=Region))  + 
  geom_point(aes(x=hyp.date, y=cpt, color=Region), alpha=0.05) + 
  geom_line(color="black", alpha=0.5) + 
  scale_y_continuous(sec.axis = sec_axis(~.*1, name="Cases Per Thousand Under 5 y.o."))+
  xlab("") +
  ylab("4 wk Lag in Square Root of Total Rainfall") +
  guides(color = guide_legend(override.aes = list(alpha=1))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

rainCPT + facet_wrap(vars(mal$Region))

tempCPT <- ggplot(data = mal, aes(x=hyp.date, y=lag16tavg, color=Region))  + 
  geom_point(aes(x=hyp.date, y=cpt, color=Region), alpha=0.1) + geom_line(color="black", alpha=0.5) + 
  scale_y_continuous(sec.axis = sec_axis(~.*1, name="Cases Per Thousand Under 5 y.o.")) + xlab("") +
  ylab("16 wk Lag in Avg. Temp") +
  guides(color = guide_legend(override.aes = list(alpha=1)))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

tempCPT + facet_wrap(vars(mal$Region))

```

```{r correlation}
# calculating correlation between lag variables and CPT

malShort <- mal[-c(1:16),]

rainCorr <- c(cor(malShort$cpt, malShort$lag2rainT), cor(malShort$cpt, malShort$lag4rainT), cor(malShort$cpt, malShort$lag8rainT), cor(malShort$cpt, malShort$lag16rainT))

tempCorr <- c(cor(malShort$cpt, malShort$lag2tavg), cor(malShort$cpt, malShort$lag4tavg), cor(malShort$cpt, malShort$lag8tavg), cor(malShort$cpt, malShort$lag16tavg))

correlation <- rbind(rainCorr, tempCorr)
colnames(correlation) <- c("2 wk Lag", "4 wk Lag", "8 wk Lag", "16 wk Lag")
rownames(correlation) <- c("RainTot:CPT Corr", "AvgTemp:CPT Corr")
kable(correlation)

```

## 2. Which region has the most malaria?

```{r histograms}

mal$Region2 <- factor(mal$Region, levels = c("Center", "Coastal", "Northern", "Southern"))
mal$Epiyear2

histCPT <- ggplot(data=mal, aes(x=Region, y=cpt, fill=Region)) + geom_bar(stat='identity')


histCPT + facet_wrap(vars(mal$Epiyear), nrow=2) + ylab("Cases per Thousand (Under 5 y.o.)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(fill="Year")


histCPT2 <- ggplot(data=mal, aes(x=Epiyear, y=cpt, fill=as.factor(Epiyear))) + geom_bar(stat='identity')
histCPT2 + facet_wrap(vars(mal$Region), nrow = 2) + labs(fill="Year")

``` 


## 3.In which regions is rainfall high? Temperature? How much do these variables vary across the country? 
```{r}
histRain <- ggplot(data=mal, aes(x=Region, y=rainTot, fill=Region)) + geom_bar(stat='identity')+ 
  ylab("Total Rainfall Across Region") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
histRain + facet_wrap(vars(mal$Epiyear), nrow=2) + 
  scale_x_discrete(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017))

histRain2 <- ggplot(data=mal, aes(x=Epiyear, y=rainTot, fill=as.factor(Epiyear))) + geom_bar(stat='identity')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Total Rainfall Across Region") + 
  scale_x_discrete(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017)) +
  labs(fill="Year")
histRain2 + facet_wrap(vars(mal$Region), nrow = 2)




```

```{r}
histTemp <- ggplot(data=mal, aes(x=Region, y=tavg, fill=Region)) + geom_boxplot()
histTemp + facet_wrap(vars(mal$Epiyear), nrow=2) + ylab("Average Temperature (C) Across Region") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

histTemp2 <- ggplot(data=mal, aes(x=Epiyear, y=tavg, fill=as.factor(Epiyear))) + geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ylab("Average Temperature (C) Across Region") 
histTemp2 + facet_wrap(vars(mal$Region), nrow = 2)
```

## Are cases clustered in a particular area of Mozambique? How does this overlap with rainfall and temperature? (think about maps here)

```{r}
# getting cases per thousand over whole districts

malaria <- as.data.frame(tapply(mal$malaria, list(mal$DISTCODE, mal$Epiyear), sum))
colnames(malaria) <- c("mal10", "mal11", "mal12", "mal13", "mal14", "mal15", "mal16")

denom <- as.data.frame(tapply(mal$denom, list(mal$DISTCODE, mal$Epiyear), sum))
colnames(denom) <- c("denom10", "denom11", "denom12", "denom13", "denom14", "denom15", "denom16")

malDenom <-as.data.frame(cbind(malaria, denom))

cptDist <- malaria/denom*1000

```


```{r maps}
library(RColorBrewer)
library(sp)
library(maptools) 
library(lattice)
library(latticeExtra) # For layer()
library(rgdal)
library(tmap)
#install.packages("tmaptools")
library(tmaptools)

poly1 <- readShapePoly('Moz_admin2.shp', IDvar="DISTCODE")
plot(poly1)
all2 <- subset(mal, Epiyear < 2017)


rainTot <- as.data.frame(tapply(all2$rainTot, list(all2$DISTCODE, all2$Epiyear), sum))
colnames(rainTot) <- c("rain10", "rain11", "rain12", "rain13", "rain14", "rain15", "rain16")
tavg <- as.data.frame(tapply(all2$tavg, list(all2$DISTCODE, all2$Epiyear), mean))
colnames(tavg) <- c("t10", "t11", "t12", "t13", "t14", "t15", "t16")
  
allStats <- as.data.frame(cbind(cptDist, rainTot, tavg))

polydat <- SpatialPolygonsDataFrame(poly1, allStats)

moz10 <- tm_shape(polydat) + 
  tm_polygons(col = "rain10" , palette="Blues", title="Total Rainfall") +
  tm_bubbles(size = "mal10", col = "t10", title.size = "Malaria Cases per Thousand", title.col = "Avg. Temp.") +
  tm_legend(outside = TRUE)

moz10

moz11 <- tm_shape(polydat) + 
  tm_polygons(col = "rain11" , palette="Blues", title="Total Rainfall") +
  tm_bubbles(size = "mal11", col = "t11", title.size = "Malaria Cases per Thousand", title.col = "Avg. Temp.") +
  tm_legend(outside = TRUE)

moz11

moz12 <- tm_shape(polydat) + 
  tm_polygons(col = "rain12" , palette="Blues", title="Total Rainfall") +
  tm_bubbles(size = "mal12", col = "t12", title.size = "Malaria Cases per Thousand", title.col = "Avg. Temp.") +
  tm_legend(outside = TRUE)
moz12

moz13 <- tm_shape(polydat) + 
  tm_polygons(col = "rain13" , palette="Blues", title="Total Rainfall") +
  tm_bubbles(size = "mal13", col = "t13", title.size = "Malaria Cases per Thousand", title.col = "Avg. Temp.") +
  tm_legend(outside = TRUE)
moz13

moz14 <- tm_shape(polydat) + 
  tm_polygons(col = "rain14" , palette="Blues", title="Total Rainfall") +
  tm_bubbles(size = "mal14", col = "t14", title.size = "Malaria Cases per Thousand", title.col = "Avg. Temp.") +
  tm_legend(outside = TRUE)
moz14

moz15 <- tm_shape(polydat) + 
  tm_polygons(col = "rain15" , palette="Blues", title="Total Rainfall") +
  tm_bubbles(size = "mal15", col = "t15", title.size = "Malaria Cases per Thousand", title.col = "Avg. Temp.") +
  tm_legend(outside = TRUE)
moz15

moz16 <- tm_shape(polydat) + 
  tm_polygons(col = "rain16" , palette="Blues", title="Total Rainfall") +
  tm_bubbles(size = "mal16", col = "t16", title.size = "Malaria Cases per Thousand", title.col = "Avg. Temp.") +
  tm_legend(outside = TRUE)
moz16


```

```{r cptDif}
poly2 <- readShapePoly('mozambique_admin1.shp', IDvar="NAME1")

mal2 <- subset(mal, mal$Province !="MAPUTO CIDADE")
# setting up data for region-based dataframe and map
#Create a new variable that is cases of malaria per thousand people #
mal2$cases <- (mal2$malaria / (mal2$Population_UN*mal2$u5weight)
)*1000

totCases <- as.data.frame(tapply(mal2$malaria, list(mal2$Province, mal2$Epiyear), sum))

colnames(totCases) <- c("cpt10", "cpt11", "cpt12", "cpt13", "cpt14", "cpt15", "cpt16")

rainTot <- as.data.frame(tapply(mal2$rainTot, list(mal2$Province, mal2$Epiyear), sum))
colnames(rainTot) <- c("rain10", "rain11", "rain12", "rain13", "rain14", "rain15", "rain16")
#tavg <- as.data.frame(tapply(mozData.2$tavg, list(mozData.2$Province, mozData.2$Epiyear), mean))
#colnames(tavg) <- c("t10", "t11", "t12", "t13", "t14", "t15", "t16")
  
allStats2 <- as.data.frame(cbind(totCases, rainTot))


#removing Maputo Cidade row from allStats
allStats2 <- allStats2[-6,]

# need to rename allStats.2 row names to match shape file
row.names(allStats2) <- c("Cabo Delgado", "Gaza", "Inhambane", "Manica", "Maputo", "Nampula", "Nassa", "Sofala", "Tete", "Zambezia")

polydat2 <- SpatialPolygonsDataFrame(poly2, allStats2)


allStats$cptDiff <- allStats$cpt16-allStats$cpt10 # difference in cases

allStats$cptDistDiff <- allStats$mal16-allStats$mal10 # difference in cpt

polydat <- SpatialPolygonsDataFrame(poly1, allStats)

# creating color palette of reds, greens, yellows, so that green is for negative (good) values,
# and red is for positive (bad) values
colors <- c("#1A9641", "#A6D96A", "#FFFFBF", "#FDAE61", "#D7191C")


cptDifMap <- tm_shape(polydat) +
  tm_polygons(col = "cptDistDiff", palette = colors, breaks = c(-Inf, 0, 5, 10, 15, 20, Inf), 
              midpoint=NA, title = "Change in Malaria Cases per Thousand: 2010-2016", border.col = "gray50", border.alpha = .5) + 
  tm_shape(polydat2) + tm_borders(lwd = 1, col = "black", alpha = .5) +
  tm_legend(outside = TRUE) 
cptDifMap



```

